Modeling trend in temperature volatility using generalized LASSO

### 2018-05-15 nips-2018/paper.tex
Rearrangement done. Comments throughout to be addressed

%Arash
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathrsfs}
\DeclareMathOperator*{\argmin}{argmin}
\usepackage{siunitx}

@@

\usepackage{xcolor}
\newcommand{\alert}[1]{\textcolor{red}{TODO: #1}}

@@

@@

Mention ~\citep{KimKoh2009,HallacPark2017}. Also, ~\citep{Tibshirani2014,TibshiraniTaylor2011}. Find Sharpnack Tibshirani. ARCH/GARCH.

@@

\section{$\ell_1$-trend filtering for estimating variance of a time-series}

$\ell_1$-trend filtering was proposed by \citep{kim_$ell_1$_2009} as a method for estimating a smooth, time-varying trend. It is formulated as the optimization problem
% : a time-series $y_t$, $t=1,...,T$ is observed and we believe that it consists of a smooth trend and a stochastic component and the goal is to estimate the trend. The estimated trend is desired to be smooth, but at the same time, we seek small estimated residuals (stochastic component). These two objectives cannot be achieved simultaneously and so a trade-off between them should be made. In $\ell_1$-trend filtering, this trade-off is formulated as the following convex optimization problem:
\begin{equation}
\min_{\beta} \frac{1}{2} \sum_{t=1}^{T} (y_t-\beta_t)^2+\lambda \sum_{t=1}^{T-2} |\beta_t-2\beta_{t+1}+\beta_{t+2}|
\end{equation}
or equivalently:
\begin{equation}
\min_{\beta} \frac{1}{2} \lVert y-\beta \lVert_2^2+\lambda \lVert D\beta\lVert_1
\label{eq:l1tf}
\end{equation}
 where $y_t$ is an observed time-series, $\beta$ is the smooth trend, $D$ is a $T-2\times T$ matrix, and $\lambda$ is a tuning parameter which balances fidelity to the data (small errors in the first term) with a desire for smoothness. 
% The \textit{penalty matrix} is:
% \begin{equation}
% D=\begin{bmatrix}
% 1 & -2 & 1 &  &  &  &\\ 
% & 1 & -2 & 1 &  &  &\\ 
% &  & \ddots & \ddots & \ddots  &  &\\ 
% &  & & 1 & -2 & 1 &  \\ 
% &  &  &   & 1 & -2 & 1 
% \end{bmatrix}
% \label{eq:d_matrix}
% \end{equation}
%The first term in \eqref{eq:l1tf} penalizes large residuals while the second term encourages smooth estimated $\beta$.
With the penalty matrix $D$, the estimated $\beta$ will be piecewise linear. \citep{kim_$ell_1$_2009} proposed a specialized primal-dual interior point (PDIP) algorithm for solving \eqref{eq:l1tf}. From a statistical perspective, \eqref{eq:l1tf} is a constrained maximum likelihood problem with independent observations from a normal distribution with common variance, $y_t \sim \mbox{N}(\beta_t, \sigma^2)$, subject to a piecewise linear constraint on $\beta$.

% The results of applying $\ell_1$-trend filtering on the Bloomington time-series are shown in \autoref{fig:bloom_detrended}. The left panel shows the original time-series (blue) together with the estimated $\beta$ (red) and the right panel shows the residuals $y_t-\beta_t$ (blue). As it can be seen, the estimated $\beta$s are piecewise linear. The number of linear segments is determined by the value of $\lambda$. Here, we chose $\lambda=500$. The choice of $\lambda$ is a model selection problem and we will return to it in the next section ???.

\subsection{Extension to variance estimation}
\label{sec:l1tf_var}

Inspired by the $\ell_1$-trend filtering algorithm, we propose a non-parametric model for estimating the variance of a time-series. To this end, we assume that at each time step $t$, there is a hidden variable $h_t$ such that conditioned on $h_t$ the observations $y_t$ are independent normal variables with zero mean and variance $\exp(h_t)$. The negative log-likelihood of the observed data in this model is $l(y|h) \propto -\sum_{t=1}^Th_t - y_t^2e^{-h_t}$. Crucially, we assume that the hidden variables $h_t$ vary smoothly. To impose this assumption, we estimate $h_t$ by solving the penalized, negative log-likelihood:
\begin{equation}
\min_h -l(y|h)+\lambda \lVert Dh \lVert_1
\label{eq:l1tf_var}
\end{equation}
 where $D$ has the same structure as above.

\alert{explain the objective more. give the AR(1) example. Explain what you loses by this assumption (ACF,forecasting). Also explain that the covariace matrix is diagonal so it cannot capture the covariance structure. But incontrast to spatial stat literature, it does not make any assumption on estimated variances. Compare to Hallac et al and Lingren et al.}

As with~\eqref{eq:l1tf}, one can solve~\eqref{eq:l1tf_var} using the PDIP algorithm (as in, e.g., \texttt{cvxopt}~\citep{andersen_cvxopt:_2013}). First, we note that this is a generalized LASSO problem \citep{tibshirani_solution_2011}. The dual of a generalized LASSO with the objective $f(x)+\lambda \lVert Dx \lVert_1$ is: 
\begin{align}
\min_\nu&\quad f^*(-D^\top\nu) \\
\mbox{s.t.}&\quad \lVert \nu \lVert_\infty \le \lambda
\end{align}
 where $f^*(\cdot)$ is the Fenchel conjugate of $f$: $f^*(u)=\max_x u^tx-f(x)$. It is simple to show that 
\begin{equation}
f^*(\mathsf{u})=\sum_t (u_t-1)\log\frac{y_t^2}{1-u_t} + u_t-1.
\label{eq:conj}
\end{equation}

\subsection{Extension to spatio-temporal data}
\label{sec:exten}

@@

At a specific time $t$, the data is measured on a grid of points with $n_r$ rows and $n_c$ columns. Let $y_{ijt}$ denote the value of the observation at time $t$ on the $i^\text{th}$ row and $j^\text{th}$ column of the grid, and $h_{ijt}$ denote the corresponding hidden variable. We seek to impose both temporal and spatial smoothness constraints on the hidden variables. Specifically, we seek a solution for $h$ which is piecewise linear in time and piecewise constant in space \footnote{The assumption that the variance is spatially piecewise constant simplifies the computations. The justification for this assumption is that we are interested in examining the trend in the variance over time and not over space.}. This can be achieved by solving the following optimization problem:

\begin{equation}
\begin{aligned}
\min_h &\sum_{i,j,t}h_{ijt}+y_{ijt}^2e^{-h_{ijt}}\\
&+\lambda_t \sum_{i,j} \sum_{t=1}^{T-2} \big|h_{ijt}-2h_{ij(t+1)}+h_{ij(t+2)}\big|\\
&+\lambda_s \sum_{t,j} \sum_{i=1}^{n_r-1} \big|h_{ijt}-h_{(i+1)jt}\big|\\
&+\lambda_s \sum_{t,i} \sum_{j=1}^{n_c-1} \big|h_{ijt}-h_{i(j+1)t}\big|
\end{aligned}
\label{eq:l1tf_var_st}
\end{equation}

The first term in the objective is the negative log-likelihood (minus a constant term). The second term is the temporal penalty for the time-series at each location $(i,j)$. The third and fourth terms, penalize the difference between the estimated variance of two vertically and horizontally adjacent points, respectively. This penalty is a special case of the penalty used in the \textit{trend filtering on graphs }\citep{wang_trend_2016} (where the difference between the estimated values of the signal at each two nodes connected with an edge is penalized.). The optimization problem \eqref{eq:l1tf_var_st} can be written in the matrix form. Let $h$ be a vector whose first $T$ entries are $h_{11t}$ for $t=1,...,T$, its next $T$ entries are $h_{21t}$ and so on. Then the optimization problem in the matrix form is as follows:

\begin{equation}
\min_h -l(y|h)+ \Lambda^t \lVert D_{total}h \lVert_1
\label{eq:l1tf_var_st_mat}
\end{equation}

Let, $\textbf{e}_n$ denote a vector of size $n$ with all entries being equal to 1. We have: $\Lambda^t=(\lambda_t\textbf{e}_{n_t}^t|\lambda_s\textbf{e}_{n_s}^t)$, $D_{total}^t=[D_{t}^t|D_{s}^t]$, where $n_t$ and $n_s$ are the number of rows of $D_{t}$ and $D_{s}$, respectively (see below). The matrix $D_t$ is the following block-diagonal matrix and corresponds to the temporal penalty:

\begin{equation}
D_t=\begin{bmatrix}
D &  & \\ 
 & \ddots & \\
 &  & D
\end{bmatrix}
\label{eq:d_t_matrix}
\end{equation}

 where $D$ was defined in \eqref{eq:d_matrix}. The number of the diagonal blocks is equal to the grid size $n_r \times n_c$. Each row of the matrix $D_s$ corresponds to one spatial constraint in \eqref{eq:l1tf_var_st}. For example, the first $T$ rows correspond to $|h_{11t}-h_{21t}|$ for $t=1,...,T$, the next $T$ rows correspond to $|h_{11t}-h_{12t}|$, and so on.

The dual of this problem is:

\begin{equation}
\begin{aligned}
\min_\nu f^*(-D_{total}^t\mathsf{\nu}) \\
s.t. \,\,\, | \nu_i | \le \Lambda_i
\end{aligned}
\label{eq:dual}
\end{equation}

 where $f^*$ is given in \eqref{eq:conj}.

\section{Optimization}
For a spatial grid of size $n_r \times n_c$ and for $T$ time steps, we have $n_t=3n_rn_c-Tn_c-2n_rn_c$ and $n_s=n_rn_cT$. For a grid over the united states and for weekly averaged data of 10 years we have $n_r=32$, $n_c=68$, $T=521$ and so $n_t\approx 3.5\times10^6$ and $n_s \approx 1.0\times10^6$. Therefore, the size of the optimization problem \eqref{eq:dual} is about $n_t\approx 4.5\times10^6$. In each step of the PDIP algorithm, we need to solve a linear system of equations in the form $z=Ax$ where $A$ is a square matrix of size $2(n_t+n_s)$ (see \citep{boyd_convex_2004} equation 11.54). Therefore, applying the PDIP directly for solving the optimization problem \eqref{eq:dual} is infeasible.

In the next section, we develop an ADMM algorithm for solving this problem. The idea is to cast the problem \eqref{eq:dual} as a so-called \textit{consensus optimization problem} \citep{boyd_distributed_2011} and solve it by breaking the problem into smaller sub-problems using an ADMM-based algorithm. Next, we first give a brief overview of the consensus optimization problem and then explain how the problem \eqref{eq:l1tf_var_st_mat} can be solved in this framework.

@@

A general form consensus optimization problem is in the form $\min_z \sum_i f_i(z)$, $z\in\mathbb{R}^n$. We can define a set of \textit{local variables} $x_i\in \mathbb{R}^{n_i}$ such that $\sum_i f_i(z)=\sum_i f_i(x_i)$. We follow the notation of \citep{boyd_distributed_2011} closely. Let $k=\mathscr{G}(i,j)$ which means that the $j^\text{th}$ entry of $x_i$ is $z_k$ (or $(x_i)_j=z_k$) and define $\tilde{z}_i \in \mathbb{R}^{n_i}$ by $(\tilde{z}_i)_j=(x_i)_j$. Then the original unconstrained optimization problem is equivalent to the following constrained optimization problem:

\begin{equation}
\begin{aligned}
\min_{\{x_1,...,x_N  \}} \sum_i f_i(x_i)\\
 s.t. \,\,\, \tilde{z}_i=x_i \,\,\,\,\,\,
\end{aligned}
\end{equation}

Now, we can apply ADMM to the \textit{augmented Lagrangian} of this problem. This results in the following ADMM updating steps at each iteration $m$:

\begin{equation}
\begin{aligned}
& x_i^{m+1}:=\argmin_{x_i} \bigg( f_i(x_i) + (u_i^m)^t x_i + (\rho/2) \lVert x_i-\tilde{z}_i^m \lVert_2^2\bigg)\\
& z_k^{m+1}:=(1/S_k)\sum_{\mathscr{G}(i,j)=k} (x_i^{m+1})_j\\
& u_i^{m+1}:=u_i^m + \rho (x_i^{m+1}-\tilde{z}_i^{m+1})
\end{aligned}
\label{eq:ADMM_steps}
\end{equation}

Here, $S_k$ is the number of local variable entries that correspond to $z_k$, and $u_i$ are the Lagrange multipliers.

To solve the optimization problem \eqref{eq:l1tf_var_st_mat} or \eqref{eq:l1tf_var_st} using the method explained in the previous section we need to two address two questions: first, how to choose the local variables $x_i$, and second, how to solve the optimization problem for updating these variables (the first line of \eqref{eq:ADMM_steps}).

Obviously, the global variable in the problem \eqref{eq:l1tf_var_st} is $z=h$. The global variable can be index by the spatial coordinate and the temporal step. In \autoref{fig:data_cube}, $z$ is represented as a cube. We can decompose $z$ into sub-cubes. An example of this decomposition is shown in the figure by white lines. It is easy to see that with this definition of $x_i$, the objective \eqref{eq:l1tf_var_st_mat} decomposes as $\sum_i f_i(x_i)$ where $f_i(x_i)=-l(y_i|x_i)+\lVert \Lambda_i^t D_ix_i \lVert_1$.

%TODO: re-write this part. it is very confusing.

\begin{figure}[tb]
	\vskip 0.2in
	\begin{center}
		\centerline{\includegraphics[width=.5\columnwidth]{Figures/data_cube}}
		\caption{The cube represents the global variable $z$ in space and time. The sub-cubes specified by the white lines are $x_i$.}
		\label{fig:data_cube}
	\end{center}
	\vskip -0.2in
\end{figure}

@@

We solve this using the PDIP method. To this end we first need to compute the dual of the problem. It can be shown that the dual of this problem is:

%TODO: write the appendix

\begin{equation}
\begin{aligned}
\min_\nu f_i^*(-D_i^t\nu_i) \\
s.t. \,\,\, | (\nu_i)_j | \le (\Lambda_i)_j
\end{aligned}
\label{eq:dual_x_update}
\end{equation}

 where:

\begin{equation}
\begin{aligned}
f^*_i(q)&=\sum_j (q_i)_j (w_i)_j -(w_i)_j-(y_i)_j^2 e^{-(w_i)_j} - \\
& (\rho/2)((w_i)_j-(\alpha_i^m)_j)^2\\
&\alpha_i^m =\hat{z}_i^m-u_i^m\\
(w_i)_j&=\mathscr{W}\bigg(\frac{(y_i)_j^2}{\rho}\exp\bigg[\frac{1-q_j-\rho(\alpha_i^m)_j}{\rho}\bigg]\bigg)-\\
& \frac{1-q_j-\rho(\alpha_i^m)_j}{\rho}
\end{aligned}
\label{eq:conj_x_update}
\end{equation}

@@

\begin{algorithm}[tb]
	\caption{ADMM for sparse estimation of variance of spatio-temporal data}
	\label{alg:ADMM}
	\begin{algorithmic}
		\STATE {\bfseries Input:} data $y$, mapping $\mathscr{G}(i,j)$, $\rho,\lambda_t,\lambda_s$
		\STATE Initialization: $x_i^0=z^0=u_i^0=\textbf{0}$.
		\FOR{$m=1,2,...$ }
			\FOR{$i=1$ {\bfseries to} $N_{sub-cubes}$}
				\STATE compute $\nu_i$ from \eqref{eq:dual_x_update}
				\STATE compute $w_i$ from \eqref{eq:conj_x_update}
				\STATE set $x_i^m:=w_i$
			\ENDFOR
		\STATE Compute $z^m$ from  \eqref{eq:ADMM_steps}
		\STATE Compute $u_i^m$ from  \eqref{eq:ADMM_steps}
		\ENDFOR
	\end{algorithmic}
\end{algorithm}

>>>>

%Arash
\usepackage{amsmath,amsthm}
\usepackage{amssymb}
\usepackage{mathrsfs}
\DeclareMathOperator*{\argmin}{argmin}
\usepackage{siunitx}

@@

\newtheorem{theorem}{Theorem}%[section]
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}

\usepackage{gensymb}
\usepackage{xcolor}
\newcommand{\attn}[1]{\textcolor{red}{TODO: #1}}
\newcommand{\one}{\mathbf{1}}
\renewcommand{\algorithmiccomment}[1]{\hfill $\rhd$ #1}
\newcommand{\given}{\;\vert\;}
\newcommand*{\algorithmautorefname}{Algorithm}%
\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}
\DeclareMathOperator*{\prox}{\bf prox}
\DeclareMathOperator*{\sign}{sign}

@@

\attn{cut this down}

@@

Mention ~\citep{KimKoh2009,HallacPark2017}. Also,
~\citep{Tibshirani2014,TibshiraniTaylor2011}. ARCH/GARCH.

@@

\begin{itemize}
\item We propose a model for non-parametric variance estimation for a
  spatio-temporal process (~\autoref{sec:ell_1-trend-filt}).
\item We derive two algorithms to fit our estimator when applied to
  very large data (~\autoref{sec:prop-optim-meth}).
\item We illustrate our methods on a large global temperature dataset
  with the goal of tracking world-wide trends in variance as well as a
  simulation constructed to mimic these data's features
  (~\autoref{sec:empirical-evaluation}). 
\end{itemize}

\section{$\ell_1$-trend filtering for estimating variance of a
  time-series}
\label{sec:ell_1-trend-filt}

$\ell_1$-trend filtering was proposed by \citep{kim_$ell_1$_2009} as a method for estimating a smooth, time-varying trend. It is formulated as the optimization problem
% : a time-series $y_t$, $t=1,...,T$ is observed and we believe that
% it consists of a smooth trend and a stochastic component and the
% goal is to estimate the trend. The estimated trend is desired to be
% smooth, but at the same time, we seek small estimated residuals
% (stochastic component). These two objectives cannot be achieved
% simultaneously and so a trade-off between them should be made. In
% $\ell_1$-trend filtering, this trade-off is formulated as the
% following convex optimization problem: 
\begin{equation}
\min_{\beta} \frac{1}{2} \sum_{t=1}^{T} (y_t-\beta_t)^2+\lambda
\sum_{t=1}^{T-2} \left|\beta_t-2\beta_{t+1}+\beta_{t+2}\right| 
\end{equation}
or equivalently:
\begin{equation}
\min_{\beta} \frac{1}{2} \norm{ y-\beta }_2^2+\lambda \norm{ D\beta}_1
\label{eq:l1tf}
\end{equation}
 where $y_t$ is an observed time-series, $\beta$ is the smooth trend,
 $D$ is a $(T-2)\times T$ matrix, and $\lambda$ is a tuning parameter
 which balances fidelity to the data (small errors in the first term)
 with a desire for smoothness.  
% The \textit{penalty matrix} is:
% \begin{equation}
% D=\begin{bmatrix}
% 1 & -2 & 1 &  &  &  &\\ 
% & 1 & -2 & 1 &  &  &\\ 
% &  & \ddots & \ddots & \ddots  &  &\\ 
% &  & & 1 & -2 & 1 &  \\ 
% &  &  &   & 1 & -2 & 1 
% \end{bmatrix}
% \label{eq:d_matrix}
% \end{equation}
%The first term in \eqref{eq:l1tf} penalizes large residuals while the second term encourages smooth estimated $\beta$.
With the penalty matrix $D$, the estimated $\beta$ will be piecewise
linear. \citep{kim_$ell_1$_2009} proposed a specialized primal-dual
interior point (PDIP) algorithm for solving \eqref{eq:l1tf}. From a
statistical perspective, \eqref{eq:l1tf} is a constrained maximum
likelihood problem with independent observations from a normal
distribution with common variance, $y_t \sim \mbox{N}(\beta_t,
\sigma^2)$, subject to a piecewise linear constraint on $\beta$.

% The results of applying $\ell_1$-trend filtering on the Bloomington
% time-series are shown in \autoref{fig:bloom_detrended}. The left
% panel shows the original time-series (blue) together with the
% estimated $\beta$ (red) and the right panel shows the residuals
% $y_t-\beta_t$ (blue). As it can be seen, the estimated $\beta$s are
% piecewise linear. The number of linear segments is determined by the
% value of $\lambda$. Here, we chose $\lambda=500$. The choice of
% $\lambda$ is a model selection problem and we will return to it in
% the next section ???.

\subsection{Estimating the variance}
\label{sec:l1tf_var}

Inspired by the $\ell_1$-trend filtering algorithm, we propose a
non-parametric model for estimating the variance of a time-series. To
this end, we assume that at each time step $t$, there is a hidden
variable $h_t$ such that conditioned on $h_t$ the observations $y_t$
are independent normal variables with zero mean and variance
$\exp(h_t)$. The negative log-likelihood of the observed data in this
model is $l(y\given h) \propto -\sum_{t=1}^Th_t - y_t^2e^{-h_t}$. Crucially,
we assume that the hidden variables $h_t$ vary smoothly. To impose
this assumption, we estimate $h_t$ by solving the penalized, negative
log-likelihood: 
\begin{equation}
\min_h -l(y\given h)+\lambda \norm{ Dh }_1
\label{eq:l1tf_var}
\end{equation}
 where $D$ has the same structure as above.

\attn{explain the objective more. give the AR(1) example. Explain what
  you loses by this assumption (ACF,forecasting). Also explain that
  the covariace matrix is diagonal so it cannot capture the covariance
  structure. But incontrast to spatial stat literature, it does not
  make any assumption on estimated variances. Compare to Hallac et al
  and Lingren et al.}

As with~\eqref{eq:l1tf}, one can solve~\eqref{eq:l1tf_var} using the
PDIP algorithm (as in, e.g.,
\texttt{cvxopt}~\citep{andersen_cvxopt:_2013}). First, we note that
this is a generalized LASSO problem
\citep{tibshirani_solution_2011}. The dual of a generalized LASSO with
the objective $f(x)+\lambda \norm{ Dx }_1$ is:  
\begin{align}
\min_\nu&\quad f^*(-D^\top\nu) & \mbox{s.t.}&\quad \norm{ \nu }_\infty \le \lambda
\end{align}
 where $f^*(\cdot)$ is the Fenchel conjugate of $f$: $f^*(u)=\max_x
 u^\top x-f(x)$. It is simple to show that 
\begin{equation}
f^*(u)=\sum_t (u_t-1)\log\frac{y_t^2}{1-u_t} + u_t-1.
\label{eq:conj}
\end{equation}
Writing
\begin{equation}
  r_w(v,\mu_1,\mu_2):=
  \begin{bmatrix}
    \nabla f^*(-D^\top v) + D(v-\lambda \one)^\top \mu_1 -
    D(v+\lambda \one)^\top \mu_2\\
    -\mu_1(v-\lambda\one)+\mu_2(v + \lambda\one) -w^{-1}\one
  \end{bmatrix}
  =
  \begin{bmatrix}
  0\\0\end{bmatrix}
\label{eq:resid}
\end{equation}
for $w>0$, where $\mu_1$ and $\mu_2$ are dual variables for the
$\ell_\infty$ constraint, as $w\rightarrow\infty$,
the solution to this nonlinear system reduces to the KKT
conditions. Therefore, the PDIP method takes Newton steps to solve the
system for a series of increasing values of $w$. \autoref{alg:pdip}
provides the details. 
\begin{algorithm}[tb]
  \caption{PDIP for $\ell_1$ variance estimation}
  \label{alg:pdip}
  \begin{algorithmic}
    \REQUIRE $\lambda>0$, $w>0$, $\nu\leftarrow 0$, $\mu_1\leftarrow
    0$, $\mu_2\leftarrow 0$, $J\in\mathbb{Z}^+$, $\{w_k\}$
    \COMMENT{Initialization} 
    \FOR[Central path]{$k=1,2,\ldots$}
    \FOR[Newton updates]{$j=1,\ldots,J$}
    \STATE Solve $A [\Delta\nu\; \Delta\mu_1\; \Delta\mu_2]^\top=r_{w_k}$ to find the search direction.
    \STATE $\quad\quad A$ is the Jacobian of $r_w$ in \eqref{eq:resid}.
    \STATE Update $[\nu^{j+1}\; \mu^{j+1}_1\; \mu^{j+1}_2] \leftarrow
    [\nu^{j}\; \mu^{j}_1\; \mu^{j}_2] + [\Delta\nu\; \Delta\mu_1\;
    \Delta\mu_2]$. 
    \ENDFOR
   \ENDFOR
   \RETURN $h=\log\frac{y^2}{1+D^\top\nu}$ \attn{Is this right?
     Explicit form of $A$}
  \end{algorithmic}
\end{algorithm}

\subsection{Adding spatial constraints}
\label{sec:exten}

@@

At a specific time $t$, the data is measured on a grid of points with
$n_r$ rows and $n_c$ columns for a total of $S=n_r\times n_c$ spatial
locations. Let $y_{ijt}$ denote the value of the 
observation at time $t$ on the $i^\text{th}$ row and $j^\text{th}$
column of the grid, and $h_{ijt}$ denote the corresponding hidden
variable. We seek to impose both temporal and spatial smoothness
constraints on the hidden variables. Specifically, we seek a solution
for $h$ which is piecewise linear in time and piecewise constant in
space (although higher-order smoothness can be imposed with minimal
alterations to the methodology). We achieve this goal
by solving the following optimization problem: 
\begin{align}
\min_h &\sum_{i,j,t}h_{ijt}+y_{ijt}^2e^{-h_{ijt}}
+\lambda_1 \sum_{i,j} \sum_{t=1}^{T-2} \left|h_{ijt}-2h_{ij(t+1)}+h_{ij(t+2)}\right|\\
&\quad+\lambda_2 \sum_{t,j} \sum_{i=1}^{n_r-1} \left|h_{ijt}-h_{(i+1)jt}\right|
+\lambda_2 \sum_{t,i} \sum_{j=1}^{n_c-1} \left|h_{ijt}-h_{i(j+1)t}\right|
\label{eq:l1tf_var_st}
\end{align}

The first term in the objective is proportional to the negative
log-likelihood, the second is the temporal penalty for the
time-series at each location $(i,j)$, while the third and fourth,
penalize the difference between the estimated variance of two
vertically and horizontally adjacent points, respectively. The spatial
component of this
penalty is a special case of trend
filtering on graphs~\citep{wang_trend_2016} which penalizes the difference
between the estimated values of the signal on the connected
nodes. As before, we can write
\eqref{eq:l1tf_var_st} in matrix form where $h$ is an $T\times
S$ vector and $D$ is replaced by $D_{TS}$. Then, as we
have two different tuning parameters for the temporal and spatial
components, we write $\lambda =\left[\lambda_1\one_T^\top,\;
  \lambda_2\one_{S}^\top\right]^\top$ and $\Lambda =\texttt{diag}(\lambda)$ leading to 
% let $h$ be a
% vector whose first $T$ entries are $h_{11t}$ for $t=1,...,T$, its next
% $T$ entries are $h_{21t}$ and so on, then the optimization problem in
% the matrix form is as
\begin{equation}
\min_h -l(y\given h)+ \norm{\Lambda D_{TS}h }_1.
\label{eq:l1tf_var_st_mat}
\end{equation}
% where $\Lambda = \mathbf{I}_{Tn_rn_c}\left[\lambda_1\one_T^\top,\;
%   \lambda_2\one_{n_rn_c}^\top\right]$ and 
% $D_{total}^t=[D_{t}^t|D_{s}^t]$, where $n_t$ and $n_s$ are the number
% of rows of $D_{t}$ and $D_{s}$, respectively (see below). The matrix
% $D_t$ is the following block-diagonal matrix and corresponds to the
% temporal penalty: 
% \begin{equation}
% D_t=\begin{bmatrix}
% D &  & \\ 
%  & \ddots & \\
%  &  & D
% \end{bmatrix}
% \label{eq:d_t_matrix}
% \end{equation}
%  where $D$ was defined in \eqref{eq:d_matrix}. The number of the
%  diagonal blocks is equal to the grid size $n_r \times n_c$. Each row
%  of the matrix $D_s$ corresponds to one spatial constraint in
%  \eqref{eq:l1tf_var_st}. For example, the first $T$ rows correspond to
%  $|h_{11t}-h_{21t}|$ for $t=1,...,T$, the next $T$ rows correspond to
%  $|h_{11t}-h_{12t}|$, and so on. 
% \attn{Detailed discussion of the PDIP to here}
The dual of this problem is 
$\min_\nu f^*(-D_{ST}^\top\nu)$ subject to 
$| \nu_k | \le \Lambda_{kk},\; k=1,\ldots,ST$,
 where $f^*$ is given in \eqref{eq:conj}.

\attn{Add back the full form of $D$ above? It was too cumbersome
  before, but possibly necessary}

\section{Proposed optimization methods}
\label{sec:prop-optim-meth}

For a spatial grid of size $S$ and $T$ time steps, $D_{ST}$ will have
$3n_rn_c+Tn_c+2Tn_r$ rows and $ST$ columns. For a $1^\circ\times
1^\circ$ grid over the
entire northern hemisphere and daily data over 10 years, we have
$n_r=90$, $n_c=180$, $T=3650$ and so $D_{ST}$ has approximately
$10^8$ columns and $10^8$
rows. In each
step of \autoref{alg:pdip}, we need to solve a linear system of
equations in $A$ which depends on $D_{ST}^\top D_{ST}$ (see \citep{boyd_convex_2004}
equation 11.54). Therefore, 
applying the PDIP directly is infeasible for our data.\footnote{We note that this
  is a highly structured and sparse matrix, but, unlike trend
  filtering alone, it is not banded. We are unaware of general
  linear algebra techniques for inverting such matrix, despite our
  best efforts.}

In the next section, we develop two ADMM algorithms for solving this
problem efficiently. The first casts the problem as a
so-called consensus optimization problem~\citep{boyd_distributed_2011}
which solves smaller sub-problems using PDIP and then recombines the results. The
second uses proximal methods to avoid matrix inversions. 
\attn{We note that stochastic gradient descent could be used but...}

@@

General (unconstrained) optimization problems $\min_z
\sum_i f_i(z)$, $z\in\mathbb{R}^n$ can be rewritten equivalently using a
collection of local variables, $x_i\in \mathbb{R}^{n_i}$, and a
constraint. Following the notation of 
\citep{boyd_distributed_2011}, let $k=\mathscr{G}(i,j)$ which
means that the $j^\text{th}$ entry of $x_i$ is $z_k$ (or
$(x_i)_j=z_k$) and define $\tilde{z}_i \in \mathbb{R}^{n_i}$ by
$(\tilde{z}_i)_j=(x_i)_j$. Then the original unconstrained
optimization problem is equivalent to the following constrained
problem
$\min_{\{x_1,...,x_N  \}} \sum_i f_i(x_i)$ subject to $\tilde{z}_i=x_i$.
Now, we can apply ADMM to the augmented Lagrangian of this
problem: 
\begin{align}
x_i&\leftarrow \argmin_{x_i} f_i(x_i) + (u_i)^\top x_i +
     (\rho/2) \norm{ x_i-\tilde{z}_i }_2^2\\
\label{eq:ADMM_steps}
z_k&\leftarrow (1/S_k)\sum_{\mathscr{G}(i,j)=k} (x_i)_j\\
u_i&\leftarrow u_i + \rho (x_i-\tilde{z}_i)
\end{align}
Here, $S_k$ is the number of local variable entries that correspond to
$z_k$, and $u_i$ are the Lagrange multipliers.  The $x$-update is
often a simple modification as the original optimization, though over only a small
partition, the $z$-update simply averages over collections of
analogous local variables while the $u$ update enforces sharing. Both
the $x$ and $u$ updates are parallelizable over the partition of local
variables.
\attn{I don't much like this explanation or notation. Seems unclear}

To solve the optimization problem \eqref{eq:l1tf_var_st_mat} or
\eqref{eq:l1tf_var_st} using this method, we need to two address two
questions: first, how to choose the 
local variables $x_i$, and second, how to solve the optimization
problem for updating these variables (the first line of
\eqref{eq:ADMM_steps}).

In \autoref{fig:data_cube}, the global variable $z$ is represented
as a cube (using the subset of the US as an example). We decompose $z$
into sub-cubes as shown in the figure by white lines. It is easy to see
that with this definition of $x_i$, the objective
\eqref{eq:l1tf_var_st_mat} decomposes as $\sum_i f_i(x_i)$ where
$f_i(x_i)=-l(y_i\given x_i)+\norm{\Lambda_{(i)} D_{(i)}x_i}$, and
$\Lambda_{(i)}$ and $D_{(i)}$ contain the temporal and spatial
penalties corresponding to $x_i$ only.
\attn{I would like to be able to write all this in terms of $h$
  instead of $x$}

\begin{figure}[tb]
  \centering
  \includegraphics[height=.2\textheight]{Figures/data_cube}
    \caption{The cube represents the global variable $z$ in space and
      time. The sub-cubes specified by the white lines are
      $x_i$. \attn{Can we use $h_i$ here instead of $x_i$?}}
    \label{fig:data_cube}
\end{figure}

@@

We solve this using the PDIP method. To this end we first need to
compute the dual of the problem. It can be shown that the dual of this
problem is: 
%TODO: write the appendix
\begin{align}
\min_\nu &f_i^*(-D_i^t\nu_i) & s.t. &| (\nu_i)_j | \le (\Lambda_i)_j
\label{eq:dual_x_update}
\end{align}
 where:
\begin{align}
f^*_i(q)&=\sum_j (q_i)_j (w_i)_j -(w_i)_j-(y_i)_j^2 e^{-(w_i)_j} - 
(\rho/2)((w_i)_j-(\alpha_i^m)_j)^2\\
\alpha_i^m &=\hat{z}_i^m-u_i^m \label{eq:conj_x_update}\\
(w_i)_j&=\mathscr{W}\bigg(\frac{(y_i)_j^2}{\rho}\exp
         \bigg[\frac{1-q_j-\rho(\alpha_i^m)_j}{\rho}\bigg]\bigg)- 
\frac{1-q_j-\rho(\alpha_i^m)_j}{\rho}
\end{align}
\attn{I'm not quite sure what's going on here. These need to match the
  generic discussion in \eqref{eq:ADMM_steps}. Or are these just
  defining some other variables for convenience? This needs to be
  clarified and incorporated into \autoref{alg:ADMM}.}

@@

\begin{algorithm}[tb]
  \caption{ADMM for sparse estimation of variance of spatio-temporal
    data \attn{Fix this}}
  \label{alg:ADMM}
  \begin{algorithmic}
    \STATE {\bfseries Input:} data $y$, mapping $\mathscr{G}(i,j)$,
    $\rho,\lambda_t,\lambda_s$ 
    \STATE Initialization: $x_i^0=z^0=u_i^0=\textbf{0}$.
    \FOR{$m=1,2,...$ }
    \FOR{$i=1$ {\bfseries to} $N_{sub-cubes}$}
    \STATE compute $\nu_i$ from \eqref{eq:dual_x_update}
    \STATE compute $w_i$ from \eqref{eq:conj_x_update}
    \STATE set $x_i^m:=w_i$
    \ENDFOR
    \STATE Compute $z^m$ from  \eqref{eq:ADMM_steps}
    \STATE Compute $u_i^m$ from  \eqref{eq:ADMM_steps}
    \ENDFOR
  \end{algorithmic}
\end{algorithm}

Because \autoref{alg:ADMM} breaks the large optimization into
sub-problems that can be solved independently, it is amenable to a
split-gather parallelization strategy via, e.g., the map reduce framework.
In each iteration, the
computation time will be equal to the time to solve each sub-problem
plus the time to communicate the solutions on the master processor
and perform the consensus step. Since each sub-problem is
small, with parallelization, the computation time in each iteration
will be small. In addition, our experiments with several values of
$\lambda_t$ and $\lambda_s$ showed that the algorithm converges in few
hundreds iterations. 
\attn{Need to redo this:}Solving each sub-problem on a machine with four
3.20GHz Intel i5-3470 cores takes less than 3 seconds on average, and
so for example if we assume that communication time is 10 seconds and
the algorithm converges in 300 iterations, with parallelization on
$N_{sub-cubes}$ machines, the algorithm will converge in about 1
hour. Assuming that we use $N_{sub-cubes}$ machines and that the
convergence rate of the algorithm is independent of the grid size,
this time will be independent of the grid size.
